Variational Monte Carlo on bosonic systems

### 2018-03-25 project_1/article/article.tex
Include results on interacting case

	\begin{table}
		\caption{Importance sampled simulations of $n=10$ particles in $d=3$ dimensions. Average acceptance
			ratios and standard deviations for eleven alpha values between $0.3$ and $0.7$.}
		\centering
			\begin{ruledtabular}
				\begin{tabular}{cccc}
				Time step & $\bar{\sigma}_b$ & $\bar{A}$ & $\expval{E}\left(\alpha = \frac{1}{2}\right)$ \\ \hline
				$2^{+3}$ & 0.09363 & 0.00531 & 15.000 \\
				$2^{+2}$ & 0.01549  & 0.04123 & 15.000 \\
				$2^{+1}$ & 0.00706 & 0.21604 & 15.000 \\ 
				$2^{\ 0}$ & 0.00675 & 0.51030 & 15.000 \\
				$2^{-1}$ & 0.00932 & 0.73779 & 15.000 \\
				$2^{-2}$ & 0.01578 & 0.86633 & 15.000 \\
				$2^{-3}$ & 0.02723 & 0.93305 & 15.000 \\
				$2^{-4}$ & 0.05016 & 0.96654 & 15.000 \\
				$2^{-5}$ & 0.08474 & 0.98316 & 15.000 \\
				$2^{-6}$ & 0.11555 & 0.99174 & 15.000 \\
				$2^{-7}$ & 0.17070 & 0.99612 & 15.000 \\
				\end{tabular}
			\end{ruledtabular}
			\label{tab:imp_time_step}
	\end{table}

	\autoref{tab:imp_time_step} shows the standard deviations and acceptance ratios for different time steps. The
	simulations in this tabel are from the same span $\alpha \in [0.3, 07]$ as in \autoref{tab:1D1N_analytic} and
	\autoref{tab:1D1N_numeric}. The optimum energies for $\alpha = \frac{1}{2}$ gives the same result for importance
	sampling as for the naïve way of sampling. Notice that by lowering the time step, the uncertainty starts to decrease
	and then starts to increase again. Moreover, the acceptance ratios increases as the time step decreases.

	\subsubsection{Test of Gradient descent}

	Before moving on to the interacting, more interesting systems, it would be beneficial to test the gradient descent implementation
	now that we are certain that it moves towards the optimum variational parameter $\alpha = \frac{1}{2}$.

\subsection{Interacting elliptical harmonic oscillator}
         \begin{table}
             \caption{The expected energy, the standard
             deviation, the acceptance ratio and CPU time for each $alpha$
             for the elliptical potential with the interacting wavefunction. We
             have used importance sampling with $\delta t = 0.1$ and $2^{21}$
             Monte Carlo cycles and an additional $10\%$ of that as
             thermalization steps. The hard sphere radius is set to $0.0043$ and
             a $\beta = \lambda = \sqrt{8}$.}
             \centering
             \begin{ruledtabular}
                 \begin{tabular}{ccccc}
                     $\alpha$ & $\langle  E_L\rangle$ & $\sigma_b$
                     &$A$&$t_C$ \\
                     \hline
                     0.2&35.17548&0.05399&0.98995&24.48868\\
                     0.3&27.62004&0.02311&0.98187&13.65132\\
                     0.4&24.97850&0.00827&0.97248&13.54666\\
                     0.5&24.39877&0.00030&0.96140&13.38388\\
                     0.6&24.83863&0.00604&0.94998&13.31570\\
                     0.7&25.82855&0.01059&0.93766&13.20579\\
                     0.8&27.22895&0.01436&0.92375&13.11933\\
                 \end{tabular}
             \end{ruledtabular}
             \label{tab:10_interacting}
         \end{table}

         \begin{table}
             \caption{$50$ interacting particles.}
             \centering
             \begin{ruledtabular}
                 \begin{tabular}{ccccc}
                     $\alpha$ & $\langle  E_L\rangle$ & $\sigma_b$
                     &$A$&$t_C$ \\
                     \hline
                     0.2&181.69371&0.25533&0.98756&199.95366\\
                     0.3&142.61591&0.10782&0.97849&195.80928\\
                     0.4&129.88615&0.04051&0.96862&190.44553\\
                     0.5&127.29926&0.00595&0.95749&195.10368\\
                     0.6&129.97630&0.03300&0.94595&193.22755\\
                     0.7&135.67382&0.05785&0.93330&191.96391\\
                     0.8&143.23238&0.07286&0.92061&192.72399\\
                 \end{tabular}
             \end{ruledtabular}
             \label{tab:50_interacting}
         \end{table}

         \begin{table}
             \caption{$100$ Interacting particles.}
             \centering
             \begin{ruledtabular}
                 \begin{tabular}{ccccc}
                     $\alpha$ & $\langle  E_L\rangle$ & $\sigma_b$
                     &$A$&$t_C$ \\
                     \hline
                     0.2&375.04401&0.51475&0.98527&745.21438\\
                     0.3&296.76879&0.20214&0.97534&741.04425\\
                     0.4&270.60830&0.08113&0.94522&723.06823\\
                     0.5&266.37263&0.02020&0.95314&702.71000\\
                     0.6&272.51171&0.07366&0.94112&689.12862\\
                     0.7&285.10285&0.11381&0.92895&707.48248\\
                     0.8&301.40609&0.15458&0.91648&707.86118\\
                 \end{tabular}
             \end{ruledtabular}
             \label{tab:100_interacting}
         \end{table}

>>>>

    \begin{table}
        \caption{Importance sampled simulations of $n=10$ particles in $d=3$ dimensions. Average acceptance
            ratios and standard deviations for eleven alpha values between $0.3$ and $0.7$.}
        \centering
            \begin{ruledtabular}
                \begin{tabular}{cccc}
                    $\delta t$ & $\bar{\sigma}_b$ & $\bar{A}$
                    & $\expval{E}\left(\alpha = \frac{1}{2}\right)$ \\
                    \hline
                    $2^{+3}$ & 0.09363 & 0.00531 & 15.000 \\
                    $2^{+2}$ & 0.01549  & 0.04123 & 15.000 \\
                    $2^{+1}$ & 0.00706 & 0.21604 & 15.000 \\
                    $2^{\ 0}$ & 0.00675 & 0.51030 & 15.000 \\
                    $2^{-1}$ & 0.00932 & 0.73779 & 15.000 \\
                    $2^{-2}$ & 0.01578 & 0.86633 & 15.000 \\
                    $2^{-3}$ & 0.02723 & 0.93305 & 15.000 \\
                    $2^{-4}$ & 0.05016 & 0.96654 & 15.000 \\
                    $2^{-5}$ & 0.08474 & 0.98316 & 15.000 \\
                    $2^{-6}$ & 0.11555 & 0.99174 & 15.000 \\
                    $2^{-7}$ & 0.17070 & 0.99612 & 15.000 \\
                \end{tabular}
            \end{ruledtabular}
            \label{tab:imp_time_step}
    \end{table}

    \autoref{tab:imp_time_step} shows the standard deviations and acceptance
    ratios for different time steps. The simulations in this table are from the
    same span $\alpha \in [0.3, 07]$ as in \autoref{tab:1D1N_analytic} and
    \autoref{tab:1D1N_numeric}. The optimum energies for $\alpha = \frac{1}{2}$
    gives the same result for importance sampling as for the naïve way of
    sampling. Notice that by lowering the time step, the uncertainty starts to
    decrease and then starts to increase again. Moreover, the acceptance ratios
    increases as the time step decreases.

    \subsubsection{Test of Gradient descent}
        Before moving on to the interacting, more interesting systems, it would
        be beneficial to test the gradient descent implementation now that we
        are certain that it moves towards the optimum variational parameter
        $\alpha = \frac{1}{2}$.

    \subsection{Interacting elliptical harmonic oscillator}
        We now look at the interacting, elliptical, harmonic oscillator with the
        scaled system shown in \autoref{eq:scaled_hamiltonian} and scaled single
        particle functions from \autoref{eq:scaled_spf} with the Jastrow factor.
        Due to similarities with the spherical harmonic oscillator system we
        make a guess that the true minimum of this new system should be situated
        close the minimum of the previous system. Choosing seven values for
        $\alpha \in [0.2, 0.7]$ we run importance sampling on this system for $N
        = \{10, 50, 100\}$ particles in $d = 3$ dimensions using $2^{21}$ Monte
        Carlo cycles\footnote{This turned out to be a little bit overkill due to
        the fast convergence of importance sampling.} and an additional $10\%$
        of the Monte Carlo cycles for thermalization of the system prior to
        doing any sampling. We have used $\delta t = 0.1$ with $a_{\text{ho}} =
        0.0043$ and $\lambda = \beta = \sqrt{8}$.

        We have sampled the expectation value for the local energy, the standard
        deviation using the blocking method, the acceptance ratio of the method
        and the CPU time spent for each value $\alpha$. The results for $N = 10$
        particles are shown in \autoref{tab:10_interacting}, $N = 50$ in
        \autoref{tab:50_interacting} and finally for $N = 100$ in
        \autoref{tab:100_interacting}.

        \begin{table}
            \caption{Simulation results for $N = 10$ bosonic interacting,
            elliptical, harmonic oscillators.}
            \centering
            \begin{ruledtabular}
                \begin{tabular}{ccccc}
                    $\alpha$ & $\langle  E_L\rangle$ & $\sigma_b$
                    &$A$ & $t_{\text{CPU}} [\si{\second}]$ \\
                    \hline
                    0.2&35.17548&0.05399&0.990%98995
                    &24.5\\%48868\\
                    0.3&27.62004&0.02311&0.981%98187
                    &13.6\\%65132\\
                    0.4&24.97850&0.00827&0.972%97248
                    &13.5\\%54666\\
                    0.5&24.39877&0.00030&0.961%96140
                    &13.4\\%38388\\
                    0.6&24.83863&0.00604&0.950%94998
                    &13.3\\%31570\\
                    0.7&25.82855&0.01059&0.938%93766
                    &13.2\\%20579\\
                    0.8&27.22895&0.01436&0.924%92375
                    &13.1\\%11933\\
                \end{tabular}
            \end{ruledtabular}
            \label{tab:10_interacting}
        \end{table}

        \begin{table}
            \caption{Simulation results for $N = 50$ bosonic interacting,
            elliptical, harmonic oscillators.}
            \centering
            \begin{ruledtabular}
                \begin{tabular}{ccccc}
                    $\alpha$ & $\langle  E_L\rangle$ & $\sigma_b$
                    & $A$ & $t_{\text{CPU}} [\si{\second}]$ \\
                    \hline
                    0.2&181.69371&0.25533&0.987%98756
                    &199.9\\%95366\\
                    0.3&142.61591&0.10782&0.978%97849
                    &195.8\\%80928\\
                    0.4&129.88615&0.04051&0.969%96862
                    &190.4\\%44553\\
                    0.5&127.29926&0.00595&0.957%95749
                    &195.1\\%10368\\
                    0.6&129.97630&0.03300&0.946%94595
                    &193.2\\%22755\\
                    0.7&135.67382&0.05785&0.933%93330
                    &192.0\\%191.96391\\
                    0.8&143.23238&0.07286&0.921%92061
                    &192.7\\%72399\\
                \end{tabular}
            \end{ruledtabular}
            \label{tab:50_interacting}
        \end{table}

        \begin{table}
            \caption{Simulation results for $N = 100$ bosonic interacting,
            elliptical, harmonic oscillators.}
            \centering
            \begin{ruledtabular}
                \begin{tabular}{ccccc}
                    $\alpha$ & $\langle  E_L\rangle$ & $\sigma_b$
                    &$A$ & $t_{\text{CPU}} [\si{\second}]$ \\
                    \hline
                    0.2&375.04401&0.51475&0.985%98527
                    &745.2\\%21438\\
                    0.3&296.76879&0.20214&0.975%97534
                    &741.0\\%04425\\
                    0.4&270.60830&0.08113&0.945%94522
                    &723.1\\%06823\\
                    0.5&266.37263&0.02020&0.953%95314
                    &702.8\\%71000\\
                    0.6&272.51171&0.07366&0.941%94112
                    &689.1\\%12862\\
                    0.7&285.10285&0.11381&0.929%92895
                    &707.5\\%48248\\
                    0.8&301.40609&0.15458&0.916%91648
                    &707.9\\%86118\\
                \end{tabular}
            \end{ruledtabular}
            \label{tab:100_interacting}
        \end{table}

        \subsubsection{Finding the minimum}
            Using the method of gradient descent we choose a set of six starting
            $\alpha_0 \in [0.2, 0.8]$ and try to locate a value for $\alpha$
            where the gradient goes to zero. We looked at $N = 10$ with $d = 3$,
            $\omega = 1$ and $\beta = \lambda = \sqrt{8}$. This yields the value
            for the minimum $\alpha$ to be
            \begin{align}
                \alpha = 0.49744 \pm 0.00002.
            \end{align}
            In \autoref{fig:gradient_descent_interacting} we can see how the
            conjugate gradient method ``moves'' to find the optimal value of
            $\alpha$.

            \begin{figure}
                \includegraphics[width=250px]{../data/figures/problem_f.pdf}
                \caption{In this figure we can see the convergence of different
                starting values for $\alpha$ towards the true value of $\alpha$
                that minimizes the expected energy in the interacting case.}
                \label{fig:gradient_descent_interacting}
            \end{figure}

\section{Discussion}
    \subsection{Energy per particle}
        For the non-interacting harmonic oscillator system there is no
        difference between a new ``particle'' and new dimensions
        \footnote{This is not entirely true in the case of our implementation of
        the Metropolis algorithms as we draw a single random particle and move
        all dimensions instead of iterating over all particles and all
        dimensions}. This means that the inclusion of a new particle or
        dimension yields a linear scaling in the energy. Thus if we compute the
        energy per particle for systems of varying size we expect to get the
        same energy. This is most easily seen in \autoref{fig:initial_problem_b}
        and in the expression for the exact energy \autoref{eq:exact_energy}.
        This, however, is \textit{not} the case in the interacting system due to
        the Jastrow factor. Looking at \autoref{tab:10_interacting} and
        \autoref{tab:100_interacting} we see that multiplying the expected
        energy of theformer with $10$ yields a lower number than the latter. We
        can also see this behavior in \autoref{fig:interacting}, where the
        energy per particle for $N = \{10, 50, 100\}$ have been plotted in the
        same figure.
